When Subgraph Isomorphism is Really Hard, and Why This Matters for Graph Databases

### 2017-11-01 paper/paper.tex
Bits of things

In \cref{section:filterverify} we look at some of the datasets commonly used to test graph database
systems, which the literature suggests are hard to solve. Simple experiments show that the entire
perceived difficulty of each of these datasets comes down to the widespread use of especially poor
subgraph isomorphism algorithms. Our fourth contribution is to show that the use of forward-checking
and the right variable-ordering heuristic, even without the more sophisticated recent advances in
subgraph algorithms, makes the entire graph database filter~/ verify paradigm unnecessary.  Finally,
we explain why filter~/ verify (and other recently proposed techniques, which we discuss in
\cref{section:angryranting}) cannot be beneficial even on more challenging instances except if an
extremely poor choice of subgraph isomorphism algorithm is made.

@@

Part of this paper were presented at IJCAI \citep{DBLP:conf/ijcai/McCreeshPT16}; the first three
sections have been expanded, whilst
\cref{section:regular,section:labelled,section:filterverify,section:angryranting} are new material.

@@

The experiments in this paper are performed on systems with Intel Xeon E5-4650 v2 CPUs, running
Scientific Linux release 6.7. We will be working with four subgraph isomorphism solvers: the Glasgow
solver \citep{DBLP:conf/cp/McCreeshP15,DBLP:conf/lion/KotthoffMS16}; LAD
\citep{DBLP:journals/ai/Solnon10}; VF2 \citep{DBLP:journals/pami/CordellaFSV04}; and VF3
\citep{DBLP:conf/gbrpr/CarlettiFSV17}. Each was compiled using GCC 4.9.

@@

?? Also plot CP

@@

We therefore believe it would be unlikely to cause too much astonishment if we suggested that a
better subgraph isomorphism algorithm could be dropped in as a black box replacement in graph
databases systems to improve their performance.  This is not our claim. Instead, this paper shows
that better algorithms invalidate the flawed premise underlying the entire filter~/ verify approach.
We will now demonstrate that filter~/ verify is simply a poor workaround for the kinds of deficiency
in VF2 demonstrated towards the end of the previous section.

@@

To demonstrate that a pure subgraph isomorphism approach is feasible, with no indexing or supporting
preprocessing, we look at four datasets commonly used in graph indexing evaluations
\citep{o:10.1371/journal.pone.0076911}. We do not claim that these are high-quality datasets or
that the associated queries are sensible, merely that they are widely used.

@@

Rather than the modern techniques used by Glasgow and LAD, we deliberately select a very simple
starting point: a constraint programming model implemented in Gecode 5.1.0 using only
toolkit-provided constraints and heuristics. The model uses a ``smallest domain first'' variable
ordering heuristic with tie-breaking on degree, an all-different constraint, extensional constraints
for adjacency, and labelled-degree filtering at the top of search. Even this simple approach finds
nearly every instance trivial, and no exponential behaviour is observed. As we show in
\cref{figure:biiiig-data}, the \emph{hardest} instance for the AIDS dataset requires 287 recursive
calls, and 149,268 of the instances can be solved without search.  For PDBS, the hardest problem
requires 3,161 recursive calls (which occurs 20 times, due to duplicated queries). For both PCM and
PPI, the hardest problem requires 23 recursive calls.  In other words, none of these instances are
in any way computationally hard even for a simple domain-based algorithm, even before introducing
stronger filtering, inference, or heuristics.  Furthermore it is certainly not the case that a
sequential scan is infeasible as so many filter~/ verify papers claim.  The exponential behaviour
seen by others in these instances is purely down to a bad choice of subgraph isomorphism algorithm,
and these instances are not inherently ``really hard''.

There are at least four benefits beyond simplicity towards abolishing filter~/ verify:

@@

\section{Further Lessons}\label{section:angryranting}

Unfortunately, indexing is not the only incorrect design choice being made in systems built around
subgraph isomorphism algorithms, and lessons from constraint programming can be applied to other
subgraph-related research. We illustrate this by looking at several recent papers on subgraph
isomorphism, showing how misleading conclusions are being reached due to domain-based algorithms not
being used or understood.

\subsection{Straggler Queries?}

@@

\subsection{Portfolios}

@@

To test this suggestion, we return briefly to the experiments on the datasets discussed earlier in
this section. What if we modify our solver so that it does not use ``smallest domain first'', and
does not detect domain wipeouts until attempting to branch on an empty variable?  Suddenly three of
the four datasets become extremely hard, with many instances now timing out after making hundreds of
millions of recursive calls, and the PPI dataset now requires hundreds of thousands of calls for
some queries.  Furthermore, slight changes to the input vertex ordering can now make some of these
``hard'' instances easy again.  This backs up our suspicions: the apparent success of
\citet{DBLP:conf/edbt/KatsarouNT17}'s technique is due to the ease of sometimes getting better
results out of a poor algorithm, rather than being an inherent hardness property, and it would not
help on really hard problem instances.

\subsection{Search Ordering}

We have seen, then, that domain tracking and domain-based ordering heuristics are critical, and we
have seen that these alone are enough to make indexing and other techniques irrelevant. There
is therefore no point in continued research such as that by
\citet{DBLP:conf/gbrpr/CarlettiFV15} and \citet{o:Carletti16}
into tweaking VF2's connectivity-based heuristics.

?? NVMe \citep{DBLP:conf/wise/ShenZ17}

?? In particular, we saw that VF3 \citep{DBLP:conf/gbrpr/CarlettiFSV17} performed \emph{worse} than
VF2 on harder instances.

\subsection{Benchmarking on Hard Instances}

?? VF3 milliseconds

?? \citet{DBLP:journals/tcbb/BonniciG17} compare various non-CP approaches to LAD. Benchmark sets
whose difficulties are measured between microseconds and milliseconds

\citet{DBLP:journals/heuristics/Gent98} describes the importance of hard benchmark instances:

@@

?? Claiming huge improvements from an algorithm portfolio consisting only of variations of poor
algorithms does not demonstrate a genuine improvement over the state of the art.

?? Need for more hard instances

?? Something about distance-based filtering being no use on our instances, low degree spread

>>>>

In \cref{section:filterverify} we look at some of the datasets commonly used to test graph database
systems, which the literature suggests are hard to solve. Simple experiments show that the entire
perceived difficulty of each of these datasets comes down to the widespread use of especially poor
subgraph isomorphism algorithms. Our fourth contribution is to show that the simplest constraint
programming approach, even without the more sophisticated recent advances in subgraph algorithms,
makes the entire graph database filter~/ verify paradigm unnecessary.  Finally, we explain why
filter~/ verify and other recently proposed techniques cannot be beneficial even on more challenging
instances except if an extremely poor choice of subgraph isomorphism algorithm is made.

@@

Part of this paper were presented at IJCAI \citep{DBLP:conf/ijcai/McCreeshPT16}; the first three
sections have been expanded, whilst
\cref{section:regular,section:labelled,section:filterverify} are new material.

@@

The experiments in this paper are performed on systems with Intel Xeon E5-4650 v2 CPUs and 768GBytes
RAM, running Scientific Linux release 6.7. We will be working with four subgraph isomorphism
solvers: the Glasgow solver \citep{DBLP:conf/cp/McCreeshP15,DBLP:conf/lion/KotthoffMS16}; LAD
\citep{DBLP:journals/ai/Solnon10}; VF2 \citep{DBLP:journals/pami/CordellaFSV04}; and VF3
\citep{DBLP:conf/gbrpr/CarlettiFSV17}. Each was compiled using GCC 4.9.

@@

@@

We therefore believe it would be unlikely to cause too much astonishment if we suggested that a
better subgraph isomorphism algorithm could be dropped in as a black box replacement in graph
databases systems to improve their performance.  This is not our claim. Instead, this paper shows
that better algorithms invalidate the flawed premise underlying the entire filter~/ verify approach.
We will now demonstrate empirically that filter~/ verify is simply a poor workaround for the kinds
of deficiency in VF2 demonstrated towards the end of the previous section.

@@

To show that a pure subgraph isomorphism approach is feasible, with no indexing or supporting
preprocessing, we look at four datasets commonly used in graph indexing evaluations. We do not claim
that these are high-quality datasets or that the associated queries are sensible, merely that
following the example of \citet{o:10.1371/journal.pone.0076911}, they are widely used.

@@

Rather than the modern techniques used by Glasgow and LAD, we deliberately select a very simple
starting point: a constraint programming model implemented in Gecode 5.1.0 using only
toolkit-provided constraints and heuristics. The model uses a ``smallest domain first'' variable
ordering heuristic with tie-breaking on degree, an all-different constraint, extensional constraints
for adjacency, and labelled-degree filtering at the top of search. Even this simple approach finds
nearly every instance in each of the four datasets trivial, and no exponential behaviour is
observed. As we show in \cref{figure:biiiig-data}, the \emph{hardest} instance for the AIDS dataset
requires 287 recursive calls, and 149,268 of the instances can be solved without search.  For PDBS,
the hardest problem requires 3,161 recursive calls (which occurs 20 times, due to duplicated
queries). For both PCM and PPI, the hardest problem requires 23 recursive calls.  In other words,
none of these instances are in any way computationally hard even for a simple domain-based
algorithm, even before introducing stronger filtering, inference, or heuristics.  Furthermore it is
certainly not the case that a sequential scan is infeasible as so many filter~/ verify papers claim.
The exponential behaviour seen by others in these instances is purely down to a bad choice of
subgraph isomorphism algorithm, and these instances are not inherently ``really hard''.

\subsection{Benefits of Constraint Programming over Filter / Verify}

There are at least four benefits beyond simplicity towards abolishing filter~/ verify in favour of a
purely constraint programming style approach:

@@

\subsection{Other Implications}

Unfortunately, indexing is not the only incorrect design choice being made in systems built around
subgraph isomorphism algorithms, and lessons from constraint programming can be applied to other
subgraph-related research. We illustrate this by looking at several recent papers on subgraph
isomorphism, discussing how a deeper understanding of the empirical hardness of subgraph isomorphism
could lead the reader to different conclusions than the ones presented.

@@

@@

To test this suggestion, we return briefly to the experiments on the datasets discussed earlier in
this section. What if we modify our solver so that it does not use ``smallest domain first'', and
does not detect domain wipeouts until attempting to branch on an empty variable?  Suddenly three of
the four datasets become extremely hard, with many instances now timing out after making hundreds of
millions of recursive calls, and the PPI dataset now requires hundreds of thousands of calls for
some queries.  Furthermore, slight changes to the input vertex ordering can now make some of these
``hard'' instances easy again.  This backs up our suspicions: the apparent success of
\citet{DBLP:conf/edbt/KatsarouNT17}'s technique is due to the ease of sometimes getting better
results out of a poor algorithm. Claiming huge improvements from an algorithm portfolio consisting
only of variations of poor algorithms does not demonstrate a genuine improvement over the state of
the art.

Improved search orderings for VF2 also appear in recent work by
\citet{DBLP:conf/gbrpr/CarlettiFV15} and \citet{o:Carletti16}, cumulating in VF3
\citep{DBLP:conf/gbrpr/CarlettiFSV17}, and by \citet{DBLP:conf/wise/ShenZ17}. We have seen in
\cref{section:labelled} that such an approach could give improvements, but will not bring an
algorithm that does not use domains close to the performance of one that does on harder instances.
Indeed, we saw in \cref{figure:induced} that VF3 performs \emph{worse} than VF2 on harder instances.
The importance of using hard benchmark instances was stressed by
\citet{DBLP:journals/heuristics/Gent98}:

@@

Unfortunately, this lesson has not sunk in. For example, \citet{DBLP:journals/tcbb/BonniciG17}
recently compared various non-constraint-based approaches to LAD using benchmark sets whose
difficulties are measured between microseconds and milliseconds. Their results suggest that LAD is
an order of magnitude slower than other approaches. This should not surprise us: the initialisation
costs of LAD are non-trivial due to it constructing domains, calculating neighbourhood degree
sequences, and performing all-different filtering. However, the risks of occasionally encountering
exponential performance from weaker algorithms on easy instances should now be clear.

Of course, hard random instances of the kind we describe in
\cref{section:non-induced,section:induced} are also atypical, and should not form the sole basis for
benchmarking. In particular, distance-based filtering
\citep{DBLP:conf/cp/AudemardLMGP14,DBLP:conf/cp/McCreeshP15,DBLP:conf/lion/KotthoffMS16} has little
to no effect on these instances, but is extremely beneficial on other kinds of graph, including
those from real-world applications.  These graphs also have a low degree spread and very little
structure to exploit, which could mislead experimenters as to the value of certain kinds of
inference. \citet{o:GentW95} give an example of an exam timetabling graph which contains an
unexpected ten-vertex clique, which would be extremely unlikely in a randomly generated graph with a
similar order and density.
